---
title: "GIS-04 - Geoprocessing - Complete"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction
This notebook focuses on geoprocessing - the process of wrangling geometric data.

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# tidyverse packages
library(dplyr)        # data wrangling

# spatial packages
library(nngeo)        # geoprocessing tools
library(mapview)      # preview spatial data
library(sf)           # spatial data tools

# other packages
library(here)         # file path management
```

## Load Data
First, we need to load up our data. This is review from the first session, and so we've pre-filled the code for you. We'll automatically re-project the data as well (also review):

```{r load-data}
# point data - 2012 Homicides in St. Louis
homicides <- st_read(here("data", "STL_CRIME_Murders12", "STL_CRIME_Murders12.shp")) %>%
  st_transform(crs = 26915)

# point data - 2018 Public Schools in Missouri
schools <- st_read(here("data", "MO_EDU_PublicK12", "MO_EDU_PublicK12.shp")) %>%
  st_transform(crs = 26915)

# polygon data 1 - St. Louis Voting Precincts
precincts <- st_read(here("data", "POL_WRD_2010_Prec", "POL_WRD_2010_Prec.shp"))%>%
  st_transform(crs = 26915)

# polygon data 2 - St. Louis Census Tracts
demos <- st_read(here("data", "STL_DEMOS_POP", "STL_DEMOS_POP.shp")) %>%
  st_transform(crs = 26915)
```

## Dissolving Features
In our `precincts` data, we have a variable named `WARD10`. This is the City Ward that each preinct falls within. If we wanted to map wards instead of precincts, we can modify our geometric data using `group_by()` and `summarise()`:

```{r dissolve-ward}
precincts %>%
  select(WARD10) %>%
  rename(ward = WARD10) %>%
  group_by(ward) %>%
  summarise() -> ward
```

Once these have been dissolved, we can explore them with `mapview()`:

```{r explore-ward}
mapview(ward)
```

Notice how some wards, such as Ward 4 and Ward 21 in North City, Ward 6 and Ward 7 in Downtown, and Wards 12, 15, and 23 in South City have "holes." These are common artifacts of the dissolve process that result from precincts' geometries not *perfectly* abutting each other.

The `nngeo` package has a great function `st_remove_holes()` that can be used to get rid of these:

```{r remove-ward-holes}
ward <- st_remove_holes(ward)
```

We can check out the differences with `mapview()`:

```{r check-ward}
mapview(ward)
```

We're going to repeat this process, but this time we'll combine a number of the wards into a single polygon for "North City."

```{r dissolve-northCity}
## create vector
northCity <- c(1, 2, 3, 4, 5, 18, 21, 22, 26, 27)

## subset
ward %>%
  mutate(northCity = ifelse(ward %in% northCity, TRUE, FALSE)) %>%
  group_by(northCity) %>%
  summarise() %>%
  filter(northCity == TRUE) %>%
  st_remove_holes() -> northCity
```

Let's inspect these data again for holes with `mapview()`:

```{r inspect-northCity}
mapview(northCity)
```

## Intersecting Data
We can use another geoprocessing technique, intersects, to identify points that lie within our polygons. For instance, we can identify which wards homicides occur in:

```{r intersect-homicides}
## intersect
homicides <- st_intersection(homicides, ward)

## preview
head(homicides)
```

We can also use this to subset our data, which is a related technique sometimes called "clipping." For example, our schools data covers the entire State of Missouri. If we only wanted schools in the City of St. Louis, we could intersect them:

```{r intersect-schools}
## intersect
schoolsSTL <- st_intersection(schools, ward)

## preview
head(schoolsSTL)
```

## Centroids
Another related geoprocessing task is using centroids to find out if polygons lie within a larger polygon or not. We use their geometric center, called a centroid, for this task. For instance, I may want to know which census tracts lie within North City. To get a sense of how these data change, we'll start by previewing our `demos` data:

```{r}
mapview(demos)
```

We'll start by calculating the geometric center (i.e. the centroid) for all of our centroids:

```{r tract-centroids}
## calculate centroids
demoCentroids <- st_centroid(demos)

## preview
mapview(demoCentroids)
```

Notice that the attributes are totally preserved. Sometimes this is ideal, other times it is not. For this example we'll do is subset our columns so that we only have `GEOID` remaining, and then intersect our centroids with our `northCity` feature:

```{r intersect-tracts}
## subset and intersect
demoCentroids %>%
  select(GEOID) %>%
  st_intersection(., northCity) -> demoCentroids

## preview
mapview(demoCentroids)
```

There are two ways to get our `TRUE` indicator back into our data. One is a "table join," the other is to create a new column by referencing our geoprocessed data. We'll illustrate the join first:

```{r join-northCity}
## make a copy of our centroid data
demoCentroids2 <- demoCentroids

## remove geometry from centroids data
st_geometry(demoCentroids2) <- NULL

## preform table join
demos_modified <- left_join(demos, demoCentroids2, by = "GEOID")

## preview
head(demos_modified)
```

Notice that tracts not in North City have `NA` value for the `northCity` column. This isn't ideal, so we'll fill those `NA` values in with `FALSE` values:

```{r fix-FALSE}
demos_modified <- mutate(demos_modified, northCity = ifelse(is.na(northCity) == TRUE, FALSE, northCity))
```

This is all pretty verbose, so we'll illustrate a quicker way to apply these data without a table join:

```{r mutate-northCity}
demos <- mutate(demos, northCity = ifelse(GEOID %in% demoCentroids$GEOID == TRUE, TRUE, FALSE))
```

This is far more succinct code, which means it is easier to debug and trouble shoot!
